Transport map accelerated-PAIS, and application to inverse problems arising from multiscale stochastic reaction networks

### 2017-06-15 TPAIS.tex
copy of thesis chunks into paper

\documentclass[final]{siamltex}
%test change
\usepackage{cite}
\usepackage{graphicx,bbm,pstricks,soul}
\usepackage{pifont}
\usepackage{bbm,algorithmic,mdframed,placeins,multirow,booktabs,subfigure}
\usepackage[ruled,vlined,linesnumbered]{algorithm2e}

@@

@@

@@

@@

>>>>

\documentclass[final]{siamltex}
%test change
\usepackage{cite}
\usepackage{graphicx,bbm,pstricks,soul}
\usepackage{pifont}
\usepackage{bbm,algorithmic,mdframed,placeins,multirow,booktabs,subfigure}
\usepackage[ruled,vlined,linesnumbered]{algorithm2e}
\usepackage{tikz,hhline}
\usepackage{tabularx}
\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{R}[1]{>{\raggedleft\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}

@@

\newcommand\irregularcircle[2]{% radius, irregularity
  \pgfextra {\pgfmathsetmacro\len{(#1)+rand*(#2)}}
  +(0:\len pt)
  \foreach \a in {10,20,...,350}{
    \pgfextra {\pgfmathsetmacro\len{(#1)+rand*(#2)}}
    -- +(\a:\len pt)
  } -- cycle
}

\newcommand{\into}{\operatornamewithlimits{\longrightarrow}}
\newtheorem{dfn}{Definition}[section]

\newcommand{\ltri}{%
\,\resizebox{!}{0.25\baselineskip}{%
\begin{tikzpicture}%
\draw[line width=1mm](0,0) -- (0,2) -- (2,0)  -- (0,0);
\end{tikzpicture}%  
}\xspace%
}%

\newcommand{\smallltri}{%
\,\resizebox{!}{0.15\baselineskip}{%
\begin{tikzpicture}%
\draw[line width=1mm](0,0) -- (0,2) -- (2,0)  -- (0,0);
\end{tikzpicture}%  
}\xspace%
}%

@@

In this section we give details on how \cite{parno2014transport} propose to couple the target,
$\mu_{\theta}$, with the reference distribution, $\mu_r$. When sampling, ideally we would like to draw a sample directly from the target distribution. When we have the exact map, $T\in\mathcal{T}$, $T\colon \mathcal{X}\rightarrow\mathbb{R}^d$, we are able to map samples from the posterior distribution onto a reference density, $\pi_r$ which is a standard Gaussian. We can then draw a proposal sample from $\pi_r$, before mapping these samples back onto target space using $T^{-1}$. These proposed samples are distributed according to the target distribution. On target space, the proposal distribution is induced by the pullback of the standard Gaussian through the map,
\begin{equation}\label{eq:pullback}
	\tilde{\pi}(\theta) = \pi_r(T(\theta))|J_T(\theta)|,
\end{equation}
which is equal to the target density when $T$ is exact. Clearly we require $T$ to be monotonic and have continuous first derivatives, which means that our space of feasible transference plans, $\mathcal{T}^\uparrow$, may not include the exact map. Since it is unlikely that $T\in\mathcal{T}^\uparrow$, we are motivated to formulate an optimisation problem as discussed in Section~\ref{sec:intro_optim}.

We attempt to find a deterministic coupling of two continuous probability distributions, $(\mu_\theta, \mu_r)$, such that $\mu_r = T\mu_\theta$ is satisfied and also that $T \in \mathcal{T}^\uparrow$. There may be infinitely many such couplings, so we look to find the coupling which minimises the difference between the proposal density and the target distribution, i.e. the distance between $\pi(\theta)$ and $\tilde{\pi}(\theta)$. In this case, we measure this distance using the Kullback-Leibler divergence.

The map does not necessarily satisfy the monotonicity condition when we optimise the cost function,
\[
	C(T) = D_\text{KL}(\pi\|\tilde{\pi}),
\]
so it must be enforced in the design of each map component. To guarantee these properties we carefully choose the basis functions for each component, and we restrict the map to be lower triangular, i.e. $T \in \mathcal{T}^{\ltri}\subset\mathcal{T}^\uparrow$. This lower triangular map has the form,
\[
	T(\theta_1, \dots, \theta_n) = \begin{bmatrix} T_1(\theta_1) \\ T_2(\theta_1, \theta_2) \\ \vdots \\
		T_n(\theta_1, \dots, \theta_n) \end{bmatrix},
\]
where $T_i\colon \mathbb{R}^i \to \mathbb{R}$. We also assume that the target and reference probability densities are absolutely continuous on
$\mathbb{R}^d$ which guarantees the maps existence.

\subsection{The optimisation problem}

With these constraints in mind we formulate the optimisation problem explicitly. The cost function
is chosen to be the Kullback-Leibler divergence between the posterior density and the pullback density,
\[
	D_\text{KL}(\pi\|\tilde{\pi}) =
		\mathbb{E}_\pi\left[\log\left(\frac{\pi(\theta)}{\tilde{\pi}(\theta)}\right)\right].
\]
Given the form of the pullback in Equation~\eqref{eq:pullback}, now taken through an approximate map, the divergence becomes
\[
	D_\text{KL}(\pi\|\tilde{\pi}) = \mathbb{E}_\pi\left[\log\pi(\theta) - \log\pi_r(\tilde{T}(\theta)) -
		\log\left|J_{\tilde{T}}(\theta)\right|\right].
\]
We note the posterior density is constant in $\tilde{T}$, and so it is not necessary for us to compute it when optimising this cost function. This expression is a complicated integral with respect to the target distribution, for which the normalisation constant is unknown, however this is exactly the scenario for which we would turn to MCMC methods for a solution.

To find the best coupling, $\tilde{T} \in \mathcal{T}^{\ltri}$, we solve the optimisation problem,
\[
	\tilde{T} = \arg\min_{T \in \mathcal{T}^{\smallltri}} \mathbb{E}_\pi\left[-\log\pi_r(T(\theta)) -
		\log\left|J_T(\theta)\right|\right]
\]
which has a unique solution. However we would like to ensure that this solution is regular and monotonic. For this
reason we regularise the problem further. The optimisation problem now takes the form
\begin{equation}\label{eq:gen_map_optim}
	\tilde{T} = \arg\min_{T\in\mathcal{T}^{\smallltri}} \left[
		 \mathbb{E}_\pi\left[-\log\pi_r(T(\theta)) -
		\log\left|J_T(\theta)\right|\right] + \beta\mathbb{E}(T(\theta)- \theta)^2 \right].
\end{equation}
This parameter $\beta$ does not need to be tuned, experimentation has shown that the choice
$\beta=1$ is sufficient for most problems. The form of the penalisation term promotes maps which are
close to the identity, and hence are likely to be monotonic.

\subsection{The structure of the map}

Before we continue with the derivation of the optimisation problem, we consider the structure
of the map in more detail. The lower lower triangular structure of the map not only guarantees monotonicity, it also allows for efficient calculation of the pullback density as well as the inverse of the map, $T^{-1}$. The Jacobian of $T$ is a lower triangular matrix,
\[
	DT(\theta) = \begin{bmatrix}
		\partial_{\theta_1} T_1(\theta) & \dots & \partial_{\theta_d} T_1(\theta)\\
		\vdots & \ddots & \vdots \\
		\partial_{\theta_1} T_d(\theta) & \dots & \partial_{\theta_d} T_d(\theta)
	\end{bmatrix} = \begin{bmatrix}
		\partial_{\theta_1} T_1(\theta) & \dots & 0\\
		\vdots & \ddots & \vdots \\
		\partial_{\theta_1} T_d(\theta) & \dots & \partial_{\theta_d} T_d(\theta)
	\end{bmatrix}
\]
since $\partial_{\theta_n} T_k(\theta) = 0$ for all $n > k$. This lower triangular structure means that the determinant of the Jacobian is a product of the diagonal elements which, when we take logs, becomes
\begin{equation}\label{eqn:separable_jacobian}
	\log\left|J_T(\theta)\right| = \sum\limits_{i=1}^d \! \log \partial_{\theta_i} T_i(\theta).
\end{equation}
Here we note that this term is separable in terms of the dimension $i$.

Inverting $T$ is simplified by the lower triangular structure in the sense that each dimension, $i$, we need only invert a univariate polynomial. This simplification comes about because we invert each dimension of the map independently, beginning with $T_1$ which is a function of only $\theta_1$. Finding the root of $T_1(\theta_1) = r_1$, gives us the value of $\theta_1$ which we can then use to find the root of $T_2(\theta_2;\theta_1) = r_2$, and so on. These roots, $r_1,\dots,r_d$, are restricted to be real since $\mathcal{X} \subset \mathbb{R}^d$, which allows us to find the unique point, $\theta$, which corresponds to the point $r$.

We require that the first derivatives of the map are continuous, which is easy to enforce by the choice of basis functions. Here we assume that the map will be built from a family of orthogonal polynomials, $\mathcal{P}(\theta)$, although not necessarily orthogonal with respect to the target distribution. Each component of the map is defined as a multivariate polynomial expansion,
\begin{equation}\label{eq:map_defn}
	\tilde{T}_i(\theta; \gamma_i) = \sum\limits_{\mathbf{j}\in\mathcal{J}_i} \!
\gamma_{i,\mathbf{j}}\psi_\mathbf{j}(\theta).
\end{equation}
The parameter $\gamma_i$ is a vector of coefficients in $\mathbb{R}^{M_i}$. Each component of $\gamma_i$ corresponds to a basis function
$\psi_\mathbf{j}$, indexed by the multi-index $\mathbf{j} \in \mathbb{N}_0^d$, here $d$ is the dimension of
$\theta$. These multi-indices are elements of the multi-index set $\mathcal{J}_i$. A multi-index defines a product of univariate polynomials in $\theta_k$,
\[
	\psi_\mathbf{j}(\theta) = \prod\limits_{k=1}^i \! \varphi_{j_k}(\theta_k), \quad \text{for} \quad \mathbf{j} \in \mathcal{J}_i,
\]
and where $\varphi_{j_k}(\theta_k) \in \mathcal{P}(\theta_k)$. Since $\tilde{T}$ is lower triangular, a multi-index $\mathbf{j}\in\mathcal{J}_i$ only contains entries for univariate polynomials in $\theta_k$ for $k\leq i$.

The cardinalities of the multi-index sets $M_i = \text{card}(\mathcal{J}_i)$ give the number of unknowns in our
optimisation problem, and so we would like to keep this number as small as possible. One option is
to use polynomials of total order $p$,
\[
	\mathcal{J}_i^\text{TO} = \left\{\mathbf{j}:\|\mathbf{j}\|_1 \leq p, j_k = 0\ \forall k > i\right\},
\]
which is optimal in terms of the amount of information captured by the map about the target. The cardinality of $\mathcal{J}_i^\text{TO}$ is $M_i = {i+p \choose p}$ which increases rapidly in $d$ and $p$. Smaller optimisation problems can be produced by constructing subsets of $\mathcal{J}_i^\text{TO}$. These index sets are discussed further
in~\cite{parno2014transport}, here we stick with polynomials of total order $p$ since we work with low dimensional problems with the PAIS algorithm.

\subsection{Implementation of the optimisation problem}\label{sec:transport_implementation}

We now return to Equation~\eqref{eq:gen_map_optim}, and show how this can be efficiently formulated
for level-2 BLAS routines~\cite{choi1992scalapack}. The Monte Carlo estimator of a function is the
sample average of the function evaluated at each sample, i.e.
\begin{align*}
	C(T) &= \mathbb{E}_\pi\left[ -\log\pi_r(T(\theta)) - \log|J_T(\theta)|\right] +
			\beta\mathbb{E}(T(\theta)-\theta)^2 \\
		&\approx \frac{1}{K}\sum\limits_{i=1}^d \! \sum\limits_{k=1}^K \left[-\log\pi_r(T_i(\theta^{(k)})) -
			\log\left|\frac{\partial T_i}{\partial \theta_i}(\theta^{(k)})\right| + \beta(T_i(\theta^{(k)})-\theta^{(k)})^2\right].
\end{align*}
This cost function will create a map from $\pi$ to any reference density $\pi_r$. We now restrict ourselves to a Gaussian reference by explicitly using a standardised Gaussian density. The above expression simplifies to
\begin{equation}\label{eq:gauss_map_optim}
	C(T) = \frac{1}{K}\sum\limits_{i=1}^d \! \sum\limits_{k=1}^K \left[\frac{1}{2}
		T_i^2(\theta^{(k)}) - \log\frac{\partial T_i}{\partial \theta_i}(\theta^{(k)}) +
		\beta(T_i(\theta^{(k)})-\theta^{(k)})^2\right].
\end{equation}
Note that since we assume that the map is monotonic, the derivatives of each component are
non-negative and so we don't need to take the absolute values in the log. In practise it is infeasible to enforce this condition across the whole parameter space. We instead enforce this condition by ensuring that the derivatives are non-negative at each sample point. This means that when we sample away from these support points while in reference space, it is possible to enter a region of space where the map is not monotonic.

We now return to the structure of the map components given in Equation~\eqref{eq:map_defn}. Since the basis functions are
fixed, the optimisation problem in \eqref{eq:gen_map_optim} is really over the map components $\bar{\gamma} = (\gamma_1, \dots,
\gamma_d)$ where $\gamma_i \in \mathbb{R}^{M_i}$. Note that $C(T)$ is the sum of $d$ expectations, and these expectations each only concern one dimension. Therefore we can rewrite \eqref{eq:gen_map_optim} as $d$ separable optimisation problems.
\begin{align}\label{eq:gamma_map_optim}
	&\arg\min_{\gamma_i\in\mathbb{R}^{M_i}} \frac{1}{K}\sum\limits_{k=1}^K
		\left[\frac{1}{2}T_i^2(\theta^{(k)}; \gamma_i) - \log\frac{\partial T_i}{\partial\theta_i}(\theta^{(k)}; \gamma_i) + \beta(T_i(\theta^{(k)};
		\gamma_i)-\theta^{(k)})^2\right], \\
	&\text{subject to} \quad \frac{\partial T_i}{\partial\theta_i}(\theta^{(k)};
		\gamma_i) > 0 \ \text{for all}\ k=1,\dots,K,\ i=1,\dots,n.
		\notag
\end{align}
The sum in Equation~\eqref{eq:map_defn} is really the inner
product between the vector of map coefficients, and the evaluations of the basis function at a
particular $\theta^{(k)}$. This means that if we organise our basis evaluations into two matrices,
$(F_i)_{k,\mathbf{j}} = \psi_\mathbf{j}(\theta^{(k)})$, and $(G_i)_{k,\mathbf{j}} =
\frac{\partial\psi_\mathbf{j}}{\partial\theta_i}(\theta^{(k)})$, for all $\mathbf{j}
\in \mathcal{J}_i^\text{TO}$, and $k = 1,\dots,K$, then we have that $T_i(\theta^{(k)}) = (F_i)_{k\cdot}\gamma_i$ and $\frac{\partial T_i}{\partial \theta_i}(\theta^{(k)}; \gamma_i) = (G_i)_{k\cdot}\gamma_i$, so \eqref{eq:gen_map_optim} becomes
\begin{align}\label{eq:blas_map_optim}
	&\arg\min_{\gamma_i\in\mathbb{R}^{M_i}} \frac{1}{2}(F_i\gamma_i)^\top(F_i\gamma_i) -
		c^\top\log(G_i\gamma_i) + \beta\sum\limits_{k=1}^K \!
		(F_i\gamma_i-\theta^{(k)})^\top(F_i\gamma_i-\theta^{(k)}), \\
	&\text{subject to} \quad G_i\gamma_i > 0. \notag
\end{align}
In this expression, the vector $c$ is a $K\times 1$ vector of ones, and $\log(G_i\gamma_i)$ is to be
evaluated element-wise. As the Monte Carlo simulations advance, new rows can be appended to the
$F_i$ and $G_i$ matrices, and $F_i^\top F_i$ can be efficiently updated via the addition of rank-1 matrices.

The regularisation term in Equation~\ref{eq:blas_map_optim} can be approximated using Parseval's identity,
\[
	\sum\limits_{k=1}^K \! (F_i\gamma_i-\theta^{(k)})^\top(F_i\gamma_i-\theta^{(k)}) \approx
		\int_{\mathbb{R}^n} |T(\theta)-\theta|^2 \text{d}\mu_\theta =
		\sum\limits_{\mathbf{j}\in\mathcal{J}_i^\text{TO}} (\gamma_{i,\mathbf{j}}-\iota_\mathbf{j})^2,
\]
where $\iota$ is the vector of coefficients for the identity map. This is of course only true when
the polynomial family $\mathcal{P}(\theta)$ is chosen to be orthonormal with respect to $\mu_\theta$; however this
approximation accomplishes the goal of efficiently regularising the optimisation problem.

These simplifications result in the efficiently implementable, regularised optimisation problem for
computing the map coefficients,
\begin{align}\label{eq:final_map_optim}
	&\arg\min_{\gamma_i\in\mathbb{R}^{M_i}} \frac{1}{2K}\gamma_i^\top F_i^\top F_i\gamma_i -
		\frac{c^\top}{K}\log(G_i\gamma_i) + \beta\|\gamma_i-\iota\|^2, \\
	&\text{subject to} \quad G_i\gamma_i > 0. \notag
\end{align}
This optimisation problem can be efficiently solved using Newton iterations. It is suggested
in~\cite{parno2014transport} that this method usually converges in around 10-15 iterations, and we
have seen no evidence that this is not a reasonable estimate. When calculating the map several times
during a Monte Carlo run, using previous guesses of the optimal map to seed the Newton algorithm
results in much faster convergence, usually taking only a couple of iterations to satisfy the stopping
criteria.

\subsection{Implementation of the optimisation problem in PAIS}

In the PAIS algorithm, we use weighted samples to approximate the posterior rather than equally weighted samples. Fortunately, the majority of the derivation of a weighted optimisation problem
follows unchanged. We look at the importance sampling Monte Carlo estimate of $C(T)$, compared with
Equation~\eqref{eq:gauss_map_optim},
\[
	C(T) = \frac{1}{\bar{w}}\sum\limits_{i=1}^n \! \sum\limits_{k=1}^K
		w_k\left[\frac{1}{2}T_i^2(\theta^{(k)}) - \log\frac{\partial
		T_i}{\partial\theta_i}(\theta^{(k)}) + \beta(T_i(\theta^{(k)})-\theta^{(k)})^2\right],
\]
here $w_k$ are the weights associated with each sample $\theta^{(k)}$, and $\bar{w}$ is the sum of
all these weights. This necessitates a minor alteration to the optimisation problem in
Equation~\eqref{eq:final_map_optim},
\begin{align}\label{eq:weighted_map_optim}
	&\arg\min_{\gamma_i\in\mathbb{R}^{M_i}} \frac{1}{2\bar{w}}\gamma_i^\top F_i^\top WF_i\gamma_i -
		\frac{w^\top}{\bar{w}}\log(G_i\gamma_i) + \beta\|\gamma_i-\iota\|^2, \\
	&\text{subject to} \quad G_i\gamma_i > 0. \notag
\end{align}
We introduce a diagonal matrix $W = \text{diag}(w)$, where $w = (w_1,\dots,w_K)^\top$ into the log of the reference density, and replace the ones vector, $c$, with $w$, otherwise the cost function is unchanged. We should obtain the same optimal map with a weighted sample as we do with an unweighted sample since we are trying to approximate the same expectation. The weights are strictly
positive resulting in a positive definite matrix $W$. This means that the Hessian of the objective
function is still positive definite, and so the optimisation problem remains convex. Again this
problem can be efficiently solved using the Newton optimisation algorithm. The Hessian takes the
form
\[
	HC_i(\gamma_i) = \frac{1}{\bar{w}}\left[F_i^\top WF_i + G_i^\top
		W\text{diag}([G_i\gamma_i]^{-2})G_i\right] + \beta I,
\]
where $[G_i\gamma_i]^{-2}$ is to be taken element-wise, and $I$ is the $M_i\times M_i$
identity matrix. The first derivative of $C_i(T)$ is
\[
	\nabla C_i(\gamma_i) = \frac{1}{\bar{w}}\left[F_i^\top WF_i\gamma_i - G_i^\top
		W[G_i\gamma_i]^{-1}\right] + \beta(\gamma_i - \iota),
\]
again $[G_i\gamma_i]^{-1}$ is taken element-wise.

@@

In a similar way, we can use the Transport map derived in Equation~\eqref{eq:weighted_map_optim} to
design a proposal scheme for the PAIS algorithm. In this case we have a choice in how to proceed; we
can either use the reference space only to propose new values, or we can run the algorithm on reference space, including resampling, and only map samples onto the target space to calculate expectations. The first option allows us to reuse much of the framework
from the standard PAIS algorithm and in the numerics later we see that this performs better than
both the Transport MH algorithm, and the standard PAIS algorithm. The second option requires some
restructuring but results in improved resampling, especially in higher dimensions.

\begin{table}
\begin{algorithm}[H]
\DontPrintSemicolon
\BlankLine
Initialise state $\theta^{(1)}_i = \theta_0$, \quad $i = 1,\dots,M$.\;
Initialise map $\bar{\gamma}^{(1)} = \iota$.\;
\For{$k \leftarrow 1, \dots, L-1$}{
	Compute $r_i = \tilde{T}(\theta^{(k)}_i; \bar{\gamma}^{(k)})$, \quad $i = 1,\dots,M$.\;
	Sample $r'_i \sim q_r(\cdot; r_i)$.\;
	Invert $\hat{\theta}_i^{(k)} = \tilde{T}^{-1}(r'_i; \bar{\gamma}^{(k)})$.\;
	Calculate:
	\[
		w_i^{(k)} = \frac{\pi(\hat{\theta}_i^{(k)})}{\left(\sum_{j=1}^M \! q_r(r_i'; r_j)\right)|J_{\tilde{T}}(\hat{\theta}_i^{(k)};\bar{\gamma}^{(k)})|}.
	\]

	Resample $\theta^{(k+1)} \leftarrow \|w^{(k)}\|^{-1}\sum\limits_{j=1}^M \! w_j^{(k)}\delta_{\hat{\theta}^{(k)}_j}(\cdot)$.\;

	\eIf{$k\ \text{mod}\ K_U = 0$ and $k < K_\text{stop}$}{
		\For{$i \leftarrow 1, \dots, n$}{
			Solve \eqref{eq:weighted_map_optim} with $\{(w^{(1)},\hat{\theta}^{(1)}), \dots, (w^{(k+1)},\hat{\theta}^{(k+1)})\}$
				and update $\gamma_i^{(k+1)}$.\;
		}
	}{
		$\bar{\gamma}^{(k+1)} = \bar{\gamma}^{(k)}$.\;
	}
}
\caption{PAIS algorithm with adaptive transport map. Option 1.\label{alg:TransportPAIS1}}
\end{algorithm}
\end{table}

The first option is given in Algorithm~\ref{alg:TransportPAIS1}. We denote the ensembles of states in target space $\theta^{(k)} = \{\theta^{(k)}_1,\dots,\theta^{(k)}_M\}$, and the states in the reference space, $r = \{r_1,\dots,r_M\}$, where $M$ is the ensemble size. Similarly, the proposal states are denoted $r' = \{r'_1,\dots,r'_M\}$ and $(w^{(k)}, \hat{\theta}^{(k)}) = \{(w^{(k)}_1, \hat{\theta}^{(k)}_1),\dots,(w^{(k)}_M, \hat{\theta}^{(k)}_M)\}$, where these pairs are the states which we consider to be our sample from the target distribution. As in the standard version of the PAIS algorithm we use the deterministic mixture weights.

\begin{table}
\begin{algorithm}[H]
\DontPrintSemicolon
\BlankLine
Initialise state $\theta^{(1)}_i = \theta_0$, \quad $i = 1,\dots,M$.\;
Initialise map $\bar{\gamma}^{(1)} = \iota$.\;
\For{$k \leftarrow 1, \dots, N-1$}{
	Compute $r_i = \tilde{T}(\theta^{(k)}_i; \bar{\gamma}^{(k)})$, \quad $i = 1,\dots,M$.\;
	Sample $r'_i \sim q_r(\cdot; r_i)$.\;
	Invert $\hat{\theta}_i^{(k)} = \tilde{T}^{-1}(r'_i; \bar{\gamma}^{(k)})$.\;
	Calculate:
	\[
		w_i^{(k)} = \frac{\pi(\hat{\theta}_i^{(k)})}{\left(\sum_{j=1}^M \! q_r(r_i'; r_j)\right)|J_{\tilde{T}}(\hat{\theta}_i^{(k)};\bar{\gamma}^{(k)})|}.
	\]

	Resample $r^* \leftarrow \|w^{(k)}\|^{-1}\sum\limits_{j=1}^M \! w_j^{(k)}\delta_{r'_j}(\cdot)$.\;
	Invert $\theta^{(k+1)}_i = \tilde{T}^{-1}(r^*_i)$.\;
	\eIf{$k\ \text{mod}\ K_U = 0$ and $k < K_\text{stop}$}{
		\For{$i \leftarrow 1, \dots, n$}{
			Solve \eqref{eq:weighted_map_optim} with $\{(w^{(1)},\hat{\theta}^{(1)}), \dots, (w^{(k+1)},\hat{\theta}^{(k+1)})\}$
				and update $\gamma_i^{(k+1)}$.\;
		}
	}{
		$\bar{\gamma}^{(k+1)} = \bar{\gamma}^{(k)}$.\;
	}
}
\caption{PAIS algorithm with adaptive transport map. Option 2.\label{alg:TransportPAIS2}}
\end{algorithm}
\end{table}

The second option, Algorithm~\ref{alg:TransportPAIS2}, is similar to the first except that rather than resampling in target space we resample in reference space. In reference space the dimensions are roughly uncorrelated, and the Gaussian marginals are easy to approximate with fewer ensemble members. This means that the resampling step will be more efficient in higher dimensions without needing to increase the ensemble size as much as we have needed to in the standard PAIS algorithm.

@@

We now move onto our main application area; the modelling of discrete chemical populations and the discovery of reaction rates within these systems. Often when modelling populations of chemical or animal species, the population is large enough that we can consider continuous models, such as coupled differential equations, to describe the evolution of the populations over time. However, this approximation is not always appropriate. These continuous models, when applied to smaller populations, result in time points where there are fractional numbers of a species. In the real world there can only be an integer number of individuals, and the effect on other species in the system might be very different to the continuous model's prediction. It is instead more appropriate to model these systems with stochastic simulation algorithms (SSAs) which model population counts as non-negative integers and take account of the intrinsic noise. The most common SSA is the Gillespie algorithm proposed in 1977~\cite{gillespie1977exact}.

\subsection{Stochastic Simulation Algorithms}\label{sec:SSAs}

SSAs are a powerful tool for describing the dynamics of discrete systems, most commonly used in the description of small scale chemical systems such as in cell biology. These algorithms simulate reactions between different species. A reaction has occured whenever the populations of the different species in the system change. For example, a simple reaction where two chemicals from the same species join together to form a chemical from a different species, this is denoted
\begin{equation}\label{eqn:chem_reaction}
	X_1 + X_1 \into^{k} X_2
\end{equation}
where the variable $k$ is the rate at which the reaction occurs. These reactions are considered to occur instantaneously. By randomly sampling the time between reactions as well as which reactions occur we can simulate a potential trajectory for the population of a species during a given time interval. Methods for the simulation of these systems assume that the system is in thermal equilibrium and that the density of each species is constant across the entire system domain.

\begin{dfn}[Stochastically exact SSA]
An SSA for a quantity $S_t$ is stochastically exact if at every time point, $t$, the probability that a particular trajectory, $T_t$, takes a value $s$ is equal to the probability of the quantity $S_t$ taking that value, i.e.
\[
	\mathbb{P}(S_t = s) = \mathbb{P}(T_t = s)\quad \forall t \ \text{and} \ s.
\]
At the time $t$, the distribution of $S_t$ is denoted as $\mu_t$.
\end{dfn}
If an SSA is stochastically exact, these trajectories can be used to estimate probability distributions for the population of a species. Empirical distributions for the value of $S_t$ throughout a time interval can be produced by a sufficiently large number of simulated trajectories.

Population models can contain several sources of variability. Large scale heterogeneity such as genetic diversity and environmental factors are often ignored in deterministic modelling. On a smaller scale the population has intrinsic noise which can come from thermal fluctuations at the molecular level~\cite{szekely2014stochastic,mcadams1999sa}. This small scale variability is inherently incorporated into the model dynamics of an SSA, whereas genetic and environmental factors are not. During the time span of a model simulation, we can assume that the genetic factors are constant and, if necessary, environmental factors can be explicitly included by evolving the probabilities of which reactions occur~\cite{shahrezaei2008colored}.

We now describe some methods of approximating $\mu_t$~\cite{gillespie1991markov}. The deterministic master equation is a system of DEs which defines a chemical system, when it is possible to solve this system of differential equations (DEs) we can obtain an exact form for $\mu_t$~\cite{anderson2016product,jahnke2007solving,anderson2010product}. However analytic solutions are rare and it is common to approximate $\mu_t$ using stochastic methods, such as the Gillespie SSA~\cite{gillespie1977exact}, $\tau$-leap methods~\cite{cao2006efficient,cao2005avoiding,chatterjee2005binomial} or SDEs such as the chemical Langevin equation~\cite{gillespie2000chemical}.

%%% Gillespie SSA

The Gillespie algorithm is an example of an exact SSA. The problem is formulated as a set of all possible reactions of the type in Equation~\eqref{eqn:chem_reaction}, the choice of which reaction and when it occurs is decided by a pair of random numbers whose distribution is determined by the relative densities of the chemical species.

\begin{dfn}[Propensity function]
Given $n$ chemical species with populations
\[
	\mathbf{X}(t) = [X_1(t),\dots,X_n(t)]^\top,
\]
the likelihood of a reaction, $R_i$, occuring is determined by the \emph{propensity function} $\alpha_i(\mathbf{X}(t); \mathbf{k})$ and depends on the current state of the system and the reaction rates $\mathbf{k}$.
\end{dfn}

The propensity function also takes into account the volume of the domain, the temperature, and many other factors. These functions $\alpha_i(\mathbf{X}(t);\mathbf{k})$ are normalised to give the probabilities, $p_i(\mathbf{X}(t);\mathbf{k})$, that $R_i$ will be the next reaction to occur,
\[
	p_i(\mathbf{X}(t); \mathbf{k}) = \frac{\alpha_i(\mathbf{X}(t);\mathbf{k})}{\alpha_0(\mathbf{X}(t);\mathbf{k})}, \quad \text{where} \quad \alpha_0(\mathbf{X}(t);\mathbf{k}) = \sum_i \! \alpha_i(\mathbf{X}(t);\mathbf{k}).
\]
There is a propensity function associated with each of the $d$ possible reactions. The form of these functions is dependent on the the type of reaction which occurs. This notation can become cumbersome and so the dependence on the current state and reaction rates may be omitted. The normalisation constant $\alpha_0$ is known as the \emph{total system propensity} or \emph{total propensity}.

The Gillespie algorithm simulates a Markov jump process about the chemical population state space. The algorithm assumes that reactions occur at times which follow an exponential distribution. The rate associated with this exponential distribution is linked to the total system propensity $\alpha_0$. Which reaction occurs at each of these times depends on the distribution $\mathbf{p} = [p_1, \dots, p_d]^\top$. This assumption is the same as that in the formulation of the chemical master equation.

Simulating a trajectory by sampling every reaction in this way allows us to generate stochastically exact draws from the true distribution $(\mu_t, \forall t)$, however it can be extremely costly, especially when reactions occur on different timescales. Variations on the algorithm have been proposed which reduce the computational cost, these are described in the review \cite{gillespie2007stochastic}.

\begin{table}[!htpb]
\centering
\begin{algorithm}[H]
\DontPrintSemicolon
\BlankLine
	Choose $\mathbf{X}(0) = [x_1, \ldots, x_n]^\top$.\;
	Define $\mathbf{k} = [k_1, \ldots, k_d]^\top$.\;
	Set $j= 0,\ t_0 = 0$.\;
	\While{$t_j < \tau$}{
		Update $\alpha_i(\mathbf{X}(t_j);\mathbf{k})$, for $i=1,\ldots, d$.\;
		Set $\alpha_0 = \sum_i \alpha_i(\mathbf{X}(t_j); \mathbf{k})$ and calculate $\mathbf{p}$.\;
		Sample $\Delta t \sim \text{Exp}(\alpha_0)$ and $r(t_j+\Delta t) \sim \text{Multinomial}(1, \mathbf{p})$.\;
		Set $t_{j+1} \leftarrow t_j + \Delta t$.\;
		Update populations $\mathbf{X}(t_{j+1}) = f_{r(t_{j+1})}(\mathbf{X}(t_j))$.\;
		$j \leftarrow j+1$.\;
	}
\caption{The Gillespie Algorithm~\cite{gillespie2007stochastic}.\label{alg:gillespie}}
\end{algorithm}
\end{table}

The Gillespie algorithm is outlined in Algorithm~\ref{alg:gillespie}. At each iteration we calculate the propensities, $\alpha_i(\mathbf{X}(t_j);\mathbf{k})$, based on the current state of each of the populations. We use these propensities to calculate $\alpha_0$ as well as the probabilities, $\mathbf{p}(\mathbf{X}(t_j); \mathbf{k})$. We now generate two random numbers, the first, $r(t_j)$, is from the multinomial distribution defined by the probability vector $\mathbf{p}$, which tells us which reaction occurs next. The second, $\Delta t$, is from the exponential distribution with rate $\alpha_0$, and tells us how long until this reaction occurs. Finally we update the system's state, $\mathbf{X}(t_{j+1})$, taking into account which reaction has occured, using the transition function $f_{r(t_j)}$. These steps result in a stochastically exact trajectory from which we can determine the probability of being in any particular state.

%%% Multiscale

In real world chemical systems, reactions will occur on more than one time scale. It is often possible to isolate which reactions are occuring more frequently (the \textit{fast reactions}) and those which are occuring less frequently (the \textit{slow reactions}). In many cases we are interested in the dynamics of the slow reactions, and the full simulation of the fast reactions can be considered a waste of computational resources. In this thesis we consider two separate simplifying approaches for these multiscale systems. The first approach uses the quasi-equilibrium assumption.

\begin{dfn}[Quasi-equilibrium assumption]
	The quasi-equilibrium assumption (QEA) is the assumption that the fast reactions converge in distribution on a timescale which is negligible with respect to the rate of occurence of the slow reactions.
\end{dfn}

This assumption allows us to approximate the dynamics of the slowly changing quantities in the system by assuming that the fast quantities are in equilibrium with respect to the fast reactions in isolation. We also consider the constrained multiscale approach (CMA) described in \cite{cotter2011constrained,cotter2016constrained}. The CMA accounts for the differences in the invariant distribution of the fast species which are due to the occurence of slow reactions. When the fast and slow subsystems operate on very different timescales, the CMA and QEA produce near identical results, however when the systems are not so clearly separated the CMA provides superior accuracy. Other simulation methods such as the $\tau$-leap method introduce significant bias into the approximations.

In this section we are interested in chemical systems where we have a small number of chemical species. We assume that we have thermal equilibrium and that the chemicals are well mixed so that we have a homogeneous system. We assume that we have accurate observations of the populations of the chemicals and the times of the reactions, and we are concerned with recovering the probability distribution which describes the rates of the reactions. The posteriors arising from situations such as these often have non-Gaussian tails which, as discussed in previous chapters, can cause problems when using kernel density estimations of the posterior distribution as we do in the standard PAIS algorithm. The dynamics of these systems are also highly non-linear resulting in complex correlation structures between parameters.

%%%

\subsection{Sufficient Statistics for Rate Recovery}

Following the assumptions for the modelling of the stochastic process $\mathbf{X}(t),\ t \in (t_0, \tau]$ outlined in the previous section, we can determine sufficient statistics for the probability distributions of the model parameters $\mathbf{k}$. Here we treat the problem in the Bayesian context and so we define a posterior distribution, $\pi(\mathbf{k}|D)$, which incorporates the model likelihoods with prior information for these reaction rates. The data, $D$, is a complete trajectory of a molecular system which includes the times of each reaction as well as which reaction occurs. In practise it might not be obvious which reaction has occured just by looking at the changes in chemical populations, however we ignore this problem for now.
